---
title: "Global Patterns Analysis"
output: html_document
---

## Dataset Overview

This section explores the structure of the GlobalPatterns microbiome dataset, including the number of samples and how they are distributed across different environments.

```{r setup, message=FALSE, warning=FALSE}
# Load required libraries
library(phyloseq)
library(ggplot2)
library(patchwork)
library(vegan)
```

```{r}
library(knitr)
library(kableExtra)

# Load dataset
data("GlobalPatterns")

# Assign to a working variable
ps <- GlobalPatterns

# Create the table
sample_counts <- table(sample_data(ps)$SampleType)

# Display with kable and white background
kable(sample_counts, 
      col.names = c("Sample Type", "Count"),
      caption = "Number of Samples per Environment Type",
      format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE,
                position = "center") %>%
  row_spec(0, bold = TRUE, background = "#2E86AB", color = "white") %>%
  column_spec(1:2, background = "white", color = "black")
```
## Data Cleaning and Filtering

To improve analysis quality, we remove low-abundance taxa that may introduce noise.

```{r filtering}
# Remove low-abundance taxa (total count <= 3)
ps_filtered <- prune_taxa(taxa_sums(ps) > 3, ps)

# Check how many taxa remain
ntaxa(ps_filtered)
```

## Community Structure (PCoA)

This analysis uses Principal Coordinates Analysis (PCoA) with Bray-Curtis distance to visualize differences in microbial community composition across sample types.

```{r pcoa_plot}
# Perform ordination using Bray-Curtis distance
ordu <- ordinate(ps_filtered, method = "PCoA", distance = "bray")

# Create PCoA plot
p_A <- plot_ordination(ps_filtered, ordu, color = "SampleType") +
  geom_point(size = 4, alpha = 0.8) +
  theme_minimal() +
  labs(title = "A: Microbial Community Structure (PCoA)")

# Display plot
p_A

# Save plot
ggsave("../results/A_pcoa.png", p_A,
       width = 7, height = 5, dpi = 300)
```

## Alpha Diversity (Shannon Index)

This analysis measures within-sample diversity using the Shannon diversity index, which accounts for both richness and evenness of microbial species.

```{r diversity_plot}
# Calculate Shannon diversity
div <- estimate_richness(ps_filtered, measures = "Shannon")

# Add sample type information
div$SampleType <- sample_data(ps_filtered)$SampleType

# Create boxplot
p_B <- ggplot(div, aes(x = SampleType, y = Shannon, fill = SampleType)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "B: Alpha Diversity (Shannon Index)",
       x = "Sample Type",
       y = "Shannon Diversity Index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# Display plot
p_B

# Save plot
ggsave("../results/B_diversity.png", p_B, width = 6, height = 5, dpi = 300)
```
## Taxonomic Composition (Phylum Level)

This visualization shows the relative abundance of microbial taxa at the phylum level across different sample types.

```{r taxa_barplot}
# Transform counts to relative abundance
ps_rel <- transform_sample_counts(ps_filtered, function(x) x / sum(x))

# Optional: keep only top taxa for cleaner plot
ps_top <- prune_taxa(names(sort(taxa_sums(ps_rel), decreasing = TRUE)[1:10]), ps_rel)

# Create bar plot
p_C <- plot_bar(ps_top, fill = "Phylum") +
  facet_wrap(~SampleType, scales = "free_x") +
  theme_minimal() +
  labs(title = "C: Taxonomic Composition (Top 10 Phyla)",
       x = "Samples",
       y = "Relative Abundance") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank())

# Display plot
p_C

# Save plot
ggsave("../results/C_taxa_barplot.png", p_C, width = 10, height = 6, dpi = 300)
```
## Phylogenetic Tree 

This visualization represents the evolutionary relationships between microbial taxa in the dataset.


```{r phylogenetic_tree_chlamydiae}
library(phyloseq)

# Subset to Chlamydiae only (use your filtered dataset)
ps_chl <- subset_taxa(ps_filtered, Phylum == "Chlamydiae")

# Remove taxa with zero counts after subsetting
ps_chl <- prune_taxa(taxa_sums(ps_chl) > 0, ps_chl)

# Create phylogenetic tree plot
p_D <- plot_tree(ps_chl, 
                 color = "SampleType",
                 shape = "Family",
                 label.tips = "Genus",
                 size = "abundance",
                 plot.margin = 0.6) +
  ggtitle("D: Phylogenetic Tree (Chlamydiae)")

# Display plot
p_D

# Save plot
ggsave("../results/D_phylogenetic_tree_chlamydiae.png", p_D,
       width = 14, height = 10, dpi = 300)
```



